Drought, vegetation productivity, and land cover change in Chile

### 2024-03-12 article/paper_drought_Chile2023.qmd
update version1 article

###	Earth observation data

For water supply and demand variables, we used ERA5L [@MunozSabater2021], a reanalys dataset that provides the evolution of land variables since 1950. It has a spatial resolution of 0.1°, hourly frequency, and global coverage. We selected the variables for total precipitation, 2 meter temperature maximum and minimum, and volumetric soil water layers between 0 and 100cm of depth (layer 1 to layer 3). The data was downloaded using the Copernicus Climate Data Store (CDS) Application Program Interface (API) implemented in `{ecmfwr}` [@Hufkens2019].

To derive a proxy for vegetation productivity, we used the product MOD13A3 collection 6.1 from MODIS [@Didan2015]. It provides vegetation indices (NDVI and EVI) at 1km of spatial resolution and monthly frequency. The MOD13A3.061 and MCD12Q1.061 were retrieved from the online Data Pool, courtesy of the NASA EOSDIS Land Processes Distributed Active Archive Center (LP DAAC), USGS Earth Resources Observation and Science (EROS) Center, Sioux Falls, South Dakota, https://lpdaa.usgs.gov/tools/data-pool/.

@@

###	Weather stations

We compared the ERA5L variables for monthly mean temperature, total precipitation, and volumetric soil water content against values retrieved by weather stations. For temperature and precipitation, we used the weather network from the Ministry of Agriculture of Chile (www.agromet.com) between 2015 and 2023. We used 277 stations located throughout Chile. For soil moisture, we select a private soil network that is owned by the agricultural enterprise Garces Fruit, which has 99 stations in Central Chile, located in cherry fruit crops. The sensors are installed at 30, 60, and 90m and are the model Teros 12 from MeterGroup. To avoid the effect of irrigation on soil moisture, which ERA5L hardly captures, we used daily data for the year 2022 and the months outside the growing season, May to September.

### Validation of ERA5L variables

To account for the performance of the ERA5L climatic variables regarding the values measured by the weather stations. We selected the following metrics:

$$MAE = \frac{1}{n}\sum |{E-S}|$$
$$Bias = \frac{\sum E}{\sum S}$$
$$ubRMSE =\sqrt{\frac{\sum{ \left[ (E_i-\overline{E})-(S_i-\overline{S}) \right ] ^2}}{n}}$$

$$CC = \frac{\sum (S_i-\bar{S})(E_i-\bar{E})}{\sqrt{(Si-\bar{S})^2(E_i-\bar{E})^2}}$$

$MAE$: mean absolute error
$bias$: bias
$ubRMSE$: unbiassed root mean squared error
$CC$: coefficient of correlation
$S$: value of the variable measure by the weather station
$E$: value of the variable measure by ERA5L

## Drought Indices

@@

For the indices EDDI and SPEI that use water demand, first we have to calculate the AED. For this, we used the method of Hargreaves [@Hargreaves1994;@Hargreaves1985]:

@@

We chose the method of Hargreaves to estimate AED because of its simplicity, which only requires temperatures and extrarrestrial radiation. Also, it has been recommended over other methods when the use of several climatic variables is limited [@Vicente-Serrano2014].

@@

We derived the drought indices of water supply and demand, soil moisture from the ERA5L dataset, and vegetation from the MODIS product, all at monthly frequency.

To evaluate water demand, we chose the $EDDI$ [@Hobbins2016; @McEvoy2016] index, which uses the $AED$. We used the SPI [@Mckee1993], a drought monitoring index that the World Meteorological Organization (WMO) recommends. We calculated the SPEI, which used a balance between $P$ and $AED$, in this case, an auxiliary variable $D = P-AED$ is used. In this study, we used the $SSI$ (standardized soil moisture index at 1 m) [@Hao2013;@AghaKouchak2014], which uses soil moisture at 1m depth. Finally, for the proxy of productivity, $zcNDVI$, we used the NDVI. Before using the NDVI, it was smoothed using a locally weighted polynomial regression, following the procedure described in @Zambrano2018 and @Zambrano2016.

All the indices are multi-scalar and were calculated for time scales of 1, 3, 6, 12, 24, and 36 months, except for zcNDVI, which was calculated for 6 months. The goal is to be able to evaluate short- and long-term droughts in water demand and supply and soil moisture. This is particularly important for central Chile because it has suffered from a prolonged decrease in precipitation for more than 12 years [@Garreaud2020;@Boisier2018;@Garreaud2017].

To calculate the drought indices, first we must calculate the accumulation of the variable. In this case, for generalization purposes, we will use $V$, referring to $P$, $AED$, $D$, $NDVI$, and $SM$ (Table \ref{tab-desEOD}). We cumulated each $V$ over the time series of $n$ values, and for the time scales $s$:

@@

It corresponds to a moving window (convolution) that sums the variable for $s$ starting for the last month $n$ until the month, which could sum for $s$ months (n-s+1). Once the variable is cumulated over time for the scale $s$, we used a nonparametric approach following @Hobbins2016 to derive the drought indices. Thus, the empirically derived probabilities are obtained through an inverse normal approximation [@Abramowitz1968]. Then, we used the empirical Tukey plotting position [@Wilks2011] over $A_i$ to derive the $P(A_i)$ probabilities across a period of interest:

@@

##	LULC change for 2001-2022 and its relation with water supply and demand, and soil moisture {#sec-methods_lulc}

### Land cover macroclasess and validation

To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection 6.1 from MODIS. This product has a yearly frequency from 2001 to 2022. The IGBP defines 17 classes; from these, we regrouped into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to schrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies. Thus, we have a land cover raster time series with the ten classes for 2001 and 2023.

To validate the land cover obtained, we compare the macroclasses with the ones of a more detailed land cover map made by @Zhao2016 for Chile with samples acquired in the years 2013–2014 (LCChile). The later has a spatial resolution of 30 m and three levels of defined classes; from those, we used level 1, which fits with the macroclasses land cover. We chose the years 2013 (IGBP2013) and 2014 (IGBP2014) from land cover macrolcasses to validate with LCChile.

We follow the next procedure:

i) resampled LCChile to the spatial resolution (500m) of the land cover macroclasses using the nearest neighbor method,
ii) took a random sample of 1000 points within continental Chile and extracted the classes that fell within each point for LCChile, IGBP2013, and IGBP2014; we considered the point extracted from LCChile as the truth and the values from the other two years as predictions.
iii) calculate a confusion matrix with the classes extracted from the 1000 points for LCChile, IGBP2013, and IGBP2014. Calculate the performance metrics of accuracy and F1.

$$Accuracy = \frac{TP+TN}{TP+TN+FP+FN}=\frac{correct\,\, classifications}{all\,\, classifications}$$
$$F1=\frac{2\cdot TP}{2\cdot TP + FP +FN}$$
where $TP$ and $FN$ refer to true positive and false negative, correctly classified classes; $TN$ and $FP$ to true negative and false positive, wrongly classified classes.

### Land cover persistence mask 2001-2022

Climate, vegetation development, seasonality, and changes in vegetation type all have an impact on the time series of NDVI. In this study, we want to examine the variation in vegetation productivity across various land cover types and how water demand, water supply, and soil moisture affect it. In order to avoid changes due to a change in the land cover type that will wrongly impact NDVI, we will develop a persistence mask for land cover for 2001–2023. Thereby, we reduce an important source of variation on a regional scale.

Therefore, we generated a raster mask for IGBP MODIS per pixel using macroclasses that remain constant for at least 80% of the years (2001–2022). This enables us to identify regions where the land cover of the macroclasses has remained constant.

### Land cover trend and drought indices

We calculated the surface occupied per land cover class into the five macrozones ("Norte Grande" to "Austral") per year for 2001–2023. After that, we calculated the trend's change in surface; we used the Sen' slope [@Sen1968] based on Mann-Kendall [@Kendall1975]. This way, we obtain a matrix of trends of 5 x 5 (macrozones x land cover). The aim is to later explore if the trend in land cover classes is associated with a trend in the drought indices. For this, we will use the techniques of regresion and regularization of Lasso [@Tibshirani2010] and Ridge [@Hoerl1970]. Also, we will test random forests for this purpose [@Ho1995]. We will choose the trend of land cover surface per macroclass and macrozone as the response variable and the trend of the drought indices (SPI, SPEI, EDDI, and SSI for time scales 1, 3, 6, 12, 24, and 36 months) as the predictor variables. With this analysis, we expect to gather insights regarding whether there is a pattern of climatic influence along Chile or if what is happening in Central Chile has to do with more localized climatic conditions.

## Trend of drought indices for water demand and supply, soil moisture, and vegetation productivity

### Mann-Kendall and Sen's slope

To estimate if there are significant positive or negative trends for the drought indices, we used the non-parametric test of Mann-Kendall [@Kendall1975]. To determine the magnitude of the trend, we used Sen's slope [@Sen1968]. Some of the advantages of applying this methodology are that the Sen's slope is not affected by outliers as regular regression does, and it is a non-paramteric method that is not affected by the distribution of the data. We applied both to the six time scales from 1981 to 2023 (monthly frequency) and the indices SPI, EDDI, SPEI, and SSI. In the case of zcNDVI (six months) was for 2000 to 2023. Thus, we have 31 trends. Also, we extracted the trend aggregated by macrozone and land cover class, obtaining a table of 31x5x5 (drought indices trends x macrozone x land cover class). We will use this data in @sec-methods_lulc to analyze if there is a strong relationship between the trends of drought indices and land cover surface within continental Chile.

### Trend in vegetation productivity without land cover change

@Vicente-Serrano2021 made a global analysis of the drought's severity trend using SPI, SPEI, and the Standardized Evapotranspiration Deficit Index (SEDI; @Vicente-Serrano2018) to evaluate AED. They indicate that the increase in hydrological drought has been due to anthropogenic effects rather than climate change. This is because the global increase in AED did not explain the change in the spatial pattern of the hydrological drought. Also, they state that _"the increase in hydrological droughts has been primarily observed in regions with high water demand and land cover change"_. We will contrast this hypothesis with what is occurring in Chile. To achieve this, we will use the land cover class type that remains more than 80% of types for 2001–2022 to evaluate the trend on zcNDVI and use this as a mask where there are low changes.

##	Impact for water supply and demand, and soil moisture in vegetation productivity within land cover types

We analyze the drought indices of water demand and supply and soil moisture against vegetation to address: i) if short- or long-term time scales are most important in impacting vegetation through Chile; and ii) the strength of the correlation for the variable and the time scale. Then, we will summarize for each land cover class and macrozone. Thus, we will be able to advance in understanding how climate is affecting vegetation, considering the impact on the five macroclasses of vegetation: forest, cropland, grassland, savanna, and shrubland.

An analysis is conducted on the linear correlation between the indices SPI, SPEI, EDDI, and SSI over time periods of 1, 3, 6, 12, 24, and 36 months, and zcNDVI. The objective is to determine the impact of soil moisture and water demand and supply on vegetation productivity. We implemented a methodology similar to that of @Meroni2016 when comparing the SPI for meteorological drought to the cumulative FAPAR (Fraction of Absorbed Photosynthetically Active Radiation), which served as an indicator for vegetation productivity. A pixel-to-pixel linear correlation analysis was performed for each index. To begin, the Pearson coefficient of correlation is computed for each of the six time scales. A significant time scale is identified as the one that attains the highest correlation (p < 0.05). Subsequently, the Pearson correlation coefficient corresponding to the time scales at which the value peaked was extracted. As a result, for each index, we generated two raster maps: one containing the time scales and the other the correlation value.

@@

### Validation of ERA5L variables

The average metrics of performance of ERA5L over the 266 weather stations were in the case of monthly temperature: $ubRMSE=1.06\,°C$, $MAE=1.131\,°C$, and $CC=0.963$, showing a good agreement, low error, and low overestimation. For cumulative monthly precipitation, $MAE=28.1\,mm$, $bias=1.93$, and $CC=0.845$, showing a high correlation and a 93% bias and being overestimated by ERA5L. In the case of the 97 soil moisture stations, we averaged for the three depths (30, 60, and 90m) and then compared it with volumetric water content at 1m derived from ERA5L. For this case, we made a daily comparison, having a $CC=0.71$, $RMSE=0.174\,m^3m^{-3}$, $MAE=0.167\,m^3m^{-3}$, and $bias=1.74$. The ERA5 soil moisture overestimate is 74%, but it has a kind of good correlation. For more detailed information, consult the supplementary material (SSX).

##	LULC change for 2001-2022 and its relation with water supply and demand, and soil moisture

### Land cover macroclasess and validation

For vegetation, we obtained and use hereafter five macroclasses of land cover from IGBP MODIS: forest, shrubland, savanna, grassland, and croplands. @fig-studyArea c shows the spatial distribution of the macroclasses through Chile for the year 2022. The validation of IGBP2013 and IGBP2014 with LCChile reached near the same metrics of performance, having an accuracy of ~0.82 and a F1 score of ~0.66 (see SS1).

### Land cover persistence mask 2001-2022 {#sec-persistence_mask}

@fig-studyArea d, shows the macroclasses of land cover persistance (80%) during 2021-2022, respectively (\ref{tab-landcoverSurf}). Within continental Chile, forest is the vegetation type with the highest surface area at 135,00 $km^2$, followed by grassland (73,176 $km^2$), savanna (54,410 $km^2$), shrubland (24,959 $km^2$), and cropland (3,100 $km^2$) (Table \ref{tab-landcoverSurf}). The macrozones with major LULCC for 2001–2022 were "Centro," "Sur," and "Austral" with 36%, 31%, and 34%, respectively (@fig-studyArea and Table \ref{tab-landcoverTrend}); of its surface that changes the type of land cover. @fig-LCprop shows the summary of the proportion of surface per land cover class and macrozone, derived from the persistance mask over continental Chile.

@@

### Land cover trend and drought indices

\begin{table}[!ht]
\caption{The value of Sen's slope trend next to the time-series plot of surface per land cover class (IGBP MCD12Q1.016) for 2001–2022 through Central Chile. Values of zero indicate that there was not a significant trend. Red dots on the plots indicate the maximum and minimum values of surface.}
\label{tab-landcoverTrend}
\includegraphics[]{../output/figs/table_var_landcover_macro.png}
\end{table}

The "Norte Chico" shows an increase in barrend land of 111 $km^2 year^{-1}$ and a reduction in the class savanna of 70 $km^2 year^{-1}$. In the "Centro" and "Sur," there are changes in the Chilean matorral, with an important reduction in savanna (136 to 318 $km^2 yr^{-1}$), and an increase in shrubland and grassland. Showing a change for more dense vegetation types. It appears to be a shift in the area of cropland from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $km^2 yr^{-1}$) in the "Sur," replacing the savanna lost.

Further, we want to address whether the trend in land cover change for 2001–2023 is associated with trends in drought indices of water demand and supply and/or soil moisture for macrozone and land cover macroclasses. From the three methods tested, Ridge, Lasso, and Random Forest, neither gives significant results regarding whether the trend in a drought index for any time scale explains the trend in land cover change. Nevertheless, in "Norte Chico" and "Centro," there is a decrease in croplands and savanna and an increase in barren land, which is associated with the variation in drought indices. Mainly for a decrease in water supply (SPI and SSI) and an increase in water demand (EDDI). However, due to the high variability from north to south in Chile, the climatic conditions (arid, semi-arid, and humid), and the land cover type, we believe that only in those zones could the LULCC be driven to some degree by drought.

##	Trend of drought indices for water demand and supply, soil moisture, and vegetation productivity

In "Norte Grande," vegetation productivity, as per the z-index, exhibits a yearly increase of 0.02 in the five land cover macroclasses, with respect to grassland and shrubland categories. There is a negative trend in "Norte Chico" with -0.04 and "Centro" with -0.02 per decade. In the "Norte Chico," savanna (-0.05) has the lowest trend, and the rest of the types are around -0.04. In "Centro," shrubland reaches -0.06, grassland -0.05, and croplands and savanna -0.01 per decade. This could be associated either with a reduction in vegetation surface, a decrease in biomass, or browning [@Miranda2023]. Vegetation reached its lowest values since the year 2019, reaching an extreme condition in early 2020 and 2022 in the "Norte Chico" and Centro" (Mega Drought). The "Sur" and "Austral" show a positive trend of around 0.016 per decade (@fig-zcNDVI_var). Despite the croplands suffering from drought just as badly as the native vegetation in "Norte Chico," the Chilean matorral [@Fuentes2021] appears to be the region most affected by a negative trend in vegetation productivity.

@@

Analyzing the water supply, the macrozones that have the lowest trend are "Norte Chico" and "Centro," where the SPI, SPEI, and SSI show that it decreases at longer time scales due to the prolonged reduction in precipitation. At 36 months, it reaches trends between -0.03 and -0.04 (z-score) per decade for SPI, SPEI, and SSI (@fig-trendDI). For "Sur," the behavior is similar, decreasing at longer scales and having between -0.016 and -0.025 per decade for SPI, SPEI, and SSI. On the other hand, all macrozones show an increase in the trend in all the drought indices, with "Norte Grande" having the highest at 36 months (0.042 per decade). Because of this, the SPEI (which uses AED) reached its lowest value in "Norte Grande," with -0.03 at 36 months. Despite the other macrozones, "Austral" showed an increase in all indices, being the highest for EDDI at 36 months (0.025) and the lowest for SSI, which shows only a minor increase in the trend (@fig-trendDI and @fig-trendDIMacro).

##	Impact for water supply and demand, and soil moisture in vegetation productivity

According to what is shown in @fig-corTimeScale, @fig-corPerson, and Table \ref{tab-corlandcover}, forest seems to be the most resistant type to drought. Showing that only "Centro" is slightly (rsq = 0.25) impacted by a 12-month soil moisture deficit (SSI-12). In the "Norte Chico" and to a lesser extent in the "Norte Grande," it is evident that a SSI-12 with a rsq = 0.45 and a decrease in water supply (SPI-36 and SPEI-24 with rsq = 0.28 and 0.34, respectively) have an impact on grasslands. However, this type was unaffected by soil moisture, water supply, or demand in macrozones further south. The types that show to be most affected by variation in climate conditions are shrublands, savannas, and croplands. For savannas in "Norte Chico," the SSI-12 and SPI-24 reached an rsq of 0.74 and 0.58, respectively. This value decreases to the south, but the SSI-12 is still the variable explaining more of the variation in vegetation productivity (rsq = 0.45 in "Centro" and 0.2 in "Sur"). In the case of croplands, the SPEI-12, SPI-36, and SSI-12 explain between 45% and 66% of "Norte Chico." The type of land most impacted by climatic variation was shrubland, where soil moisture explained 59% and precipitation, 37%, in "Norte Chico" and "Centro," with SSI-12 being the most relevant variable, then SPI-36 in "Norte Chico" and SPI-24 in "Sur."

@@

>>>>

### Gridded meteorological and vegetation data

To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection 6.1 from MODIS. This product has a yearly frequency from 2001 to 2022. The IGBP defines 17 classes; from these, we regrouped into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies. Thus, we have a land cover raster time series with the ten classes for 2001 and 2023. Prior to using this class, we compared it with the types of land cover made by @Zhao2016, which is a more detailed land cover map of Chile with a 30 m spatial resolution for 2012–2013. It reached a global accuracy of ~0.82. The procedure of validation is demonstrated in Section S1 of the supplementary material.To derive a proxy for vegetation productivity, we used the MOD13A3 collection 6.1 product from MODIS [@Didan2015]. It provides vegetation indices (NDVI and EVI) at 1km of spatial resolution and monthly frequency. The MOD13A3 and MCD12Q1 were retrieved from the online Data Pool, courtesy of the NASA EOSDIS Land Processes Distributed Active Archive Center (LP DAAC), USGS Earth Resources Observation and Science (EROS) Center, Sioux Falls, South Dakota, https://lpdaa.usgs.gov/tools/data-pool/.

@@

For soil moisture, water supply, and water demand variables, we used ERA5L [@MunozSabater2021], a reanalysis dataset that provides the evolution of the meteorological and soil moisture variables since 1950. It has a spatial resolution of 0.1° (9 km), hourly frequency, and global coverage. We selected the variables for total precipitation, maximum and minimum temperature at 2 meters, and volumetric soil water layers between 0 and 100cm of depth (layer 1 to layer 3).

## Land cover change and trend

To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection 6.1 from MODIS. @Zambrano2018 and @Fuentes2021 have previously used this product for studies of drought and land cover. The MCD12Q1 has a yearly frequency from 2001 to 2022. The IGBP defines 17 classes; from these, we regrouped into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies. Thus, we have a land cover raster time series with the ten macroclasses for 2001 and 2023. We validate the land cover macroclasses regarding a highly detailed (30 m of spatial resolution) land cover map made for Chile by @Zhao2016 for 2013-2014. Our results showed a global accuracy of ~0.82 and a F1 score of ~0.66. Section S2 in the Supplementary Material shows the procedure for validation.

Climate, vegetation development, seasonality, and changes in vegetation type all have an impact on the time series of NDVI. In this study, we want to examine the variation in vegetation productivity across various land cover types and how water demand, water supply, and soil moisture affect it. In order to avoid changes due to a change in the land cover type that will wrongly impact NDVI, we developed a persistence mask for land cover for 2001–2022. Thereby, we reduce an important source of variation on a regional scale. Therefore, we generated a raster mask for IGBP MODIS per pixel using macroclasses that remained unchanged for at least 80% of the years (2001–2022). This enabled us to identify regions where the land cover macroclasses are persistent.
We calculated the surface occupied per land cover class into the five macrozones ("Norte Grande" to "Austral") per year for 2001–2023. After that, we calculated the trend’s change in surface per type. We used the Sen’ slope [@Sen1968] based on Mann-Kendall [@Kendall1975].

## Trend of drought and land cover change

@@

In order to compute the drought indices that uses water demand, it is necessary to first calculate the AED. To do this, we employed the Hargreaves technique [@Hargreaves1994;@Hargreaves1985], by applying the following equation:

@@

We chose the method of Hargreaves to estimate AED because of its simplicity, which only requires temperatures and extrarrestrial radiation. Also, it has been recommended over other methods (e.g., Penman-Monteith) when the access to climatic variables is limited [@Vicente-Serrano2014].

@@

To derive the drought indices of water supply and demand, soil moisture, and vegetation we used the ERA5L dataset and the MODIS product, with a monthly frequency for 1981-2023 and 2000-2023, respectively.

The dought indices correspond to a historical anomaly with regard to a variable (e.g., meteorological, vegetation, or soil moisture). To account for the anomaly, the common practice is to derive it following a statistical parametric methodology in which it is assumed that the statistical distribution of the data is known (@Heim2002). A wrong decision is usually the highest source of uncertainty (@Laimighofer2022). In the case of Chile, due to its high degree of climatic variability, it is complex to choose a proper distribution without previous research. Here, we follow a non-parametric methodology for the calculation of the drought indices, in a similar manner as the framework proposed by @Farahmand2015; @Hobbins2016;@McEvoy2016.

For the purpose of monitoring water supply drought, we used the well-known Standardized Precipitation Index (SPI), which the World Meteorological Organization (WMO) recommended. The SPI solely relies on precipitation data. Also, it has been used worldwide for the study of drought, including in Chile (@Garreaud2017; @Zambrano2017). The primary cause of drought is precipitation anomalies, and temperature usually makes it worse (Luo et al. 2017). Nowadays, there is an increase in attention toward using water demand separately to monitor droughts. (@Vicente-Serrano2020; @Noguera2022). Thus, to evaluate water demand, we chose the Evaporative Demand Drought Index (EDDI), developed by Hobbins et al. (016) and McEvoy et al. (2016), which is based on the AED. EDDI is currently used for monitoring drought in the United States (https://psl.noaa.gov/eddi/). In our case, we used only temperature for AED, a difference from the original formulation of EDDI, which also considered wind besides temperature. To consider the combined effect of water supply and demand, we selected the SPEI, which corresponds to a balance between precipitation and AED. @Vicente-Serrano2010 proposed the SPEI, and it has improved the SPI by incorporating temperature for drought monitoring. For SPEI, an auxiliary variable D = P-AED is calculated. Soil moisture is the main driver of vegetation productivity, particularly in semi-arid regions (@Li2022). Hence, for soil water drought, we used the SSI (Standardized Soil Moisture Index) (Hao and AghaKouchak 2013; A. AghaKouchak 2014) which is a multi-scale index similar to SPI, SPEI, and EDDI. In our case, for the SSI, we used the average soil moisture from ERA5L at 1m depth. Finally, for the proxy of productivity, we used the zcNDVI proposed by Zambrano et al. (2018) which will be derived from the NDVI retrieved from MOD13A1.

To derive the drought indices, first we must calculate the sum of the variables with regard to the time scale (s). In this case, for generalization purposes, we will use $V$, referring to variables $P$, $AED$, $D$, $NDVI$, and $SM$ (#ref{tab-desEOD}). We cumulated each $V$  over the time series of $n$ values (months), and for the time scales $s$:

@@

The $A_{si}$ corresponds to a moving window (convolution) that sums the variable for time scales $s$ from the last month, month by month, until the first month in which it could sum for $s$ months. Once the variable is cumulated over time for the scale $s$. Thus, the empirically derived probabilities are obtained through an inverse normal approximation [@Abramowitz1968]. Then, we used the empirical Tukey plotting position [@Wilks2011] over $A_i$ to derive the $P(a_i)$  probabilities across a period of interest:

@@

The drought indices were calculated for time scales of 1, 3, 6, 12, 24, and 36 months at a monthly frequency for 1981–2023 in order to be used for short- to long-term evaluation of drought. In the case of the proxy of vegetation productivity (zcNDVI) it was calculated for a time scale of six months at monthly frequency for 2000–2023. For zcNDVI, we test time scales of 1, 3, 6, and 12 months; we choose to use six months because that shows a more robust representation of vegetation productivity due to the seasonality of vegetation in Chile.

### Trend of drought indices

To estimate if there are significant positive or negative trends for the drought indices, we used the non-parametric test of Mann-Kendall [@Kendall1975]. To determine the magnitude of the trend, we used Sen’s slope [@Sen1968]. Some of the advantages of applying this methodology are that the Sen’s slope is not affected by outliers as regular regression does, and it is a non-parametric method that is not influenced by the distribution of the data. We applied Mann-Kendall to see if the trend was significant and Sen's slope to estimate the magnitude of the trend. We did this to the six time scales from 1981 to 2023 (monthly frequency) and the indices SPI, EDDI, SPEI, and SSI. Thus, we have trends per index and time scale (24 in total). Then, we extracted the trend aggregated by macrozone and per land cover persitent macroclasses.

## Drought impacts on vegetation productivity within land cover

We analyzed the trend of vegetation productivity over the unchanged land cover macroclasses. This way, we tried to reduce the noise in the vegetation due to a change in land cover from year to year. To achieve this, we will use the persistent mask of land cover macroclasses, which are the types that remain more than 80% of the time for 2001–2022. We used this to evaluate the trend in zcNDVI per land cover class and macrozone.

We examine the drought indices of water demand, water supply, soil moisture, and their connection with vegetation productivity to investigate two main questions: i) whether short-term or long-term time scales have a greater impact on vegetation across Chile and its specific regions; and ii) the spatial variation in the strength of the correlation between the variables and time scales.  Then, we will summarize for each land cover class and macrozone. Thus, we will be able to advance in understanding how climate is affecting vegetation, considering the impact on the five macroclasses of vegetation: forest, cropland, grassland, savanna, and shrubland.

We conducted an analysis on the linear correlation between the indices SPI, SPEI, EDDI, and SSI over time periods of 1, 3, 6, 12, 24, and 36 months, and zcNDVI. The objective is to determine the impact of soil moisture and water demand and supply on vegetation productivity. We used a method similar to that used by @Meroni2017 which compared the SPI with the cumulative FAPAR (Fraction of Absorbed Photosynthetically Active Radiation). A pixel-to-pixel linear correlation analysis was performed for each index within the persistent mask of land cover macroclasses. To begin, the Pearson coefficient of correlation is computed for each of the six time scales. A significant time scale is identified as the one that attains the highest correlation (p < 0.05). Subsequently, the Pearson correlation coefficient corresponding to the time scales at which the value peaked was extracted. As a result, for each index, we generated two raster maps: 1) containing the raster with values of the time scales that reached the maximum correlation, and 2) having the value of the correlation obtained.

@@

# Land cover change and trend

@@

For vegetation, we obtained and use hereafter five macroclasses of land cover from IGBP MODIS: forest, shrubland, savanna, grassland, and croplands. [@fig-studyArea]c shows the spatial distribution of the macroclasses through Chile for the year 2022. [@fig-studyArea]d shows the macroclasses of land cover persistance (80%) during 2021–2022, respectively (#ref{tab-landcoverSurf}). Within continental Chile, barren land is the land cover class with the highest surface area (277,870 km2). The largest type of vegetation, with 137,085 km2, is forest. Grassland (74,247 km2), savanna (55,206 km2), shrubland (25,341 km2), and cropland (3,146 km2) are the other types (Table 2). The macrozones with major changes for 2001–2022 were "Centro," "Sur," and "Austral," with 36%, 31%, and 34% of their surface changing the type of land cover, respectively (@fig-studyArea and #ref{tab-landcoverTrend}). @fig-LCprop shows the summary of the proportion of surface per land cover class and macrozone, derived from the persistance mask over continental Chile.

```{=latex}
\begin{table}[!ht]
\caption{The value of Sen's slope trend next to the time-series plot of surface per land cover class (IGBP MCD12Q1.016) for 2001–2022 through Central Chile. Values of zero indicate that there was not a significant trend. Red dots on the plots indicate the maximum and minimum values of surface.}
\label{tab-landcoverTrend}
\includegraphics[]{../output/figs/table_var_landcover_macro.png}
\end{table}
```

The "Norte Chico" shows an increase in barren land of 111 $km^2 yr^{-1}$  and a reduction in the class savanna of 70 $km^2 yr^{-1}$. In the "Centro" and "Sur," there are changes with an important reduction in savanna (136 to 318 $km^2 yr^{-1}$ ), and an increase in shrubland and grassland. Showing a change for more dense vegetation types. It appears to be a shift in the area cultivated (croplands) from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $km^2 yr^{-1}$ ) in the "Sur," replacing the savanna lost (#ref{tab-landcoverTrend}).

## Trend on drought and its relation to land cover change

### Trend of drought indices

@fig-trendDI shows the spatial variation of the trend for the drought indices from short- to long-term scales. SPI and SPEI have a decreasing trend from "Norte Chico" to "Sur." However, there is an increasing trend in "Austral." The degree of the trend is stronger at higher time scales. The SSI indicates that in "Norte Grande," there are surfaces that have increased in the soutwest part and in the northeast have decreased, and is shown for all time scales. Similar to SPI and SPEI, SSI decreases at higher time scales. EDDI showed a positive trend for the whole of continental Chile, with a higher trend toward the north and a descending gradient toward the south. The degree of trend increases at higher time scales.

@@

In @fig-trendDIMacro, the averaged aggregation per macrozone, drought index, and time scale are shown. The macrozones that have the lowest trend are "Norte Chico" and "Centro," where the SPI, SPEI, and SSI show that it decreases at longer time scales. Potentially explained due to the prolonged reduction in precipitation that has affected the hydrological system in Chile. At 36 months, it reaches trends between -0.03 and -0.04 (z-score) per decade for SPI, SPEI, and SSI. For "Sur," the behavior is similar, decreasing at longer scales and having between -0.016 and -0.025 per decade for SPI, SPEI, and SSI. "Norte Grande" has the highest trend at 36 months for EDDI (0.042 per decade), and "Centro" has the lowest for SPI and SPEI. In "Norte Grande" and "Norte Chico," which are in a semi-arid climate, it is evident that the EDDI has an effect on the difference between the SPI and SPEI index, which is not seen in the other macrozones. Contrary to the other macrozones, “Austral” showed an increase in all indices, being the highest for EDDI at 36 months (0.025) and the lowest for SSI, which shows only a minor increase in the trend.

### Relationship between drought indices and land cover change

We look at the relationship between drought index and land cover change in Figure 5 by comparing the trends in land cover change (in terms of the total surface area per land cover type and macrozone) and drought indexes. Figure 5 shows that the negative trends in cropland (-0.029) and savanna (-0.026) in “Norte Chico” are the highest and are associated with an increase trend in EDDI and a decreasing trend in SPI, SPEI, and SSI. On the contrary, the shrubland in “Centro” has an increase (0.04) linked to a decrease in SPI, SPEI, and SSI and to an increase in EDDI. In "Austral,” the positive trend in shrubland fits with the positive trend in all the drought indices. The rest of the land cover types in the macrozones show weak associations with water supply, water demand, or soil moisture drought indices.

```{r figTrendsLandDrought}
#| echo: false
#| fig-pos: "!ht"
#| fig-cap: "Relationship between the trend in land cover change and the trend in drought indices for the five macrozones. Vertical panels correspond to 1, 3, 6, 12, 24, and 36 months of the time scale by drought index. Horizontal panels show each drought index"
#| out-width: "\\textwidth"
knitr::include_graphics('../output/figs/points_landcover_drought_indices_trend_and_time_scale.png')
```

## Drought impacts on vegetation productivity within land cover

### Vegetation productivity

In @fig-zcNDVI_var it is showed the spatial map of trends in zcNDVI ([fig-zcNDVI_var]a) and the temporal variation of zcNDVI within the aggregated macrozones ([fig-zcNDVI_var]b). In "Norte Grande," vegetation productivity, as per the z-index, exhibits a yearly increase of 0.02 with respect to grassland and shrubland categories. There is a negative trend in "Norte Chico" with -0.04 and "Centro" with -0.02 per decade. In the "Norte Chico," savanna (-0.05) has the lowest trend, and the rest of the types are around -0.04. In "Centro," shrubland reaches -0.06, grassland -0.05, and croplands and savanna -0.01 per decade. This could be associated either with a reduction in vegetation surface, a decrease in biomass, or browning [@Miranda2023]. Vegetation reached its lowest values since the year 2019, with an extreme condition in early 2020 and 2022 in the "Norte Chico" and "Centro". The "Sur" and "Austral" show a positive trend of around 0.016 per decade (@fig-zcNDVI_var). Despite the croplands suffering from drought just as badly as the native vegetation in "Norte Chico," the savanna and shurbland appears to be the region most affected by a negative trend in vegetation productivity.

### Correlation between vegetation productivity and drought indices

@@

@fig-corTimeScale is a map that shows the highest coefficient of determination (r-squared, or rsq) found in the regression analysis between different drought indicators and plant productivity over time. The spatial variation of time scales reached per index is mostly for time scales above 12 months. In the case of SSI, the predominant scales are 6 and 12 months. For all indices, to the north, the time scales are higher and diminish toward the south until the south part of "Austral" increases. In @fig-corPerson, the map of Pearson correlation values is shown. The EDDI reached correlations above 0.5 between "Norte Chico" and "Sur." The correlation changes from negative to positive toward the Andes Mountains and to the sea, just as in the northern part of "Austral." The SPI and SPEI have similar results, with the higher values in "Norte Chico" and "Centro" being higher than 0.6. Following a similar spatial pattern as EDDI. The SSI showed to be the index that has a major spatial extension with a higher correlation. It has a similar correlation to SPI and SPEI for "Norte Chico" and "Sur,"  but has improvements for "Sur."
